---
title: 'STAT 630: Homework 3'
author: "Andrea Labra Orozco"
date: 'Due: September 24, 2022'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Concept Questions

The following questions relate to the `nycflights` dataset found in the `openintro` package. *Note, you may have to coerce some of the columns to factors befor your start.*

Find more information on the dataset [here](http://openintrostat.github.io/oilabs-tidy/02_intro_to_data/intro_to_data.html)

```{r, message = FALSE, warning = FALSE}
library(openintro)
library(dplyr)
library(ggplot2)
data("nycflights")
#?nycflights # Description of the dataset (comment out before knitting)
```

For as many questions as you can, practice using inline R code to display your answers. For example:

The number of flights in the dataset is $n$ = `r nrow(nycflights)`.

I will put a \* next to questions you might want to try this with.

1. Pick an airline carrier (see link above for carrier codes) and a month of the year (1-12). Input your choices into the code below. The R chunk will create a histogram of `arr_delay` (arrival delay in minutes) for the airline and month you chose. Then, using `stat_function` ggplot2 will overlay a true normal distribution with the mean and standard deviation of `arr_delay` for your carrier/month. 

```{r}
# Subset the data according to your chosen/carrier/month
my_flights <- nycflights %>% 
  filter(carrier == "DL", month == 12) 

# Calculate mean/sd of arr_delay
arr_delay_stats <- my_flights %>% 
  summarise(x_bar = mean(arr_delay),
            s = sd(arr_delay))

# Plot histogram of arr_delay, overlaying true normal curve 
my_flights %>% 
  ggplot(aes(arr_delay)) +
  geom_histogram(aes(y = ..density..)) + # relative frequency
  stat_function(fun = dnorm, # overlay a stat function (a normal dist)
                args = list(mean = arr_delay_stats$x_bar, # let mu = x_bar
                            sd = arr_delay_stats$s)) # let sigma = s
```

2. Comment on the normality of `arr_delay` compared to the true normal curve.
The histogram of the data shows a roughly bell-shaped distribution with a slight right skew, which is observable as a few bars extending further to the right. While the histogram resembles a normal curve to some extent, there is a very high peak where the normal peak is expected. However, when examining the Q-Q plot, the majority of points closely follow the reference line, indicating that the data approximates a normal distribution reasonably well. In context, this means that most of the flights are delayed at the times that are expected to be delayed .

For questions 3 & 4, assume the distribution of `arr_delay` for your chosen carrier and month is normally distributed with the mean and standard deviation found in Question 1.

3. \*Using R functions, find the probability that the flight on DL during December was early.

```{r}

# Filter the dataset for Delta Airlines (DL) flights in December
delta_december_flights <- my_flights %>% 
  filter(carrier == "DL", month == 12)

# Calculate the probability of being early (arr_delay <= 0)
#TRUE for early flights (arr_delay <= 0) and FALSE for late flights (arr_delay > 0).
probability_early <- mean(delta_december_flights$arr_delay <= 0)

# Concatenate and Print 
cat("The probability that a Delta Airlines flight in December was early is:", probability_early )


```



4. \*Using R functions, calculate the probability that a randomly selected flight on DL in December is between 30 to 60 minutes late.

```{r}

# Filter the dataset for Delta Airlines (DL) flights in December with delays between 30 and 60 minutes
delta_december_late_flights <- my_flights %>% 
  filter(carrier == "DL", month == 12, arr_delay >= 30, arr_delay <= 60)

# Calculate the total number of Delta Airlines flights in December
total_delta_december_flights <- my_flights %>% 
  filter(carrier == "DL", month == 12)

# Calculate the probability of being between 30 and 60 minutes late
probability_late <- nrow(delta_december_late_flights) / nrow(total_delta_december_flights)

cat("The probability that a randomly selected Delta Airlines flight in December is between 30 to 60 minutes late is:", probability_late)


```


5. \*Use`quantile()` to calculate the actual 95th percentile (0.95 quantile) from the `arr_delay` data on DL in December. Interpret this value in the context of the problem.
```{r}
# Filter the dataset for Delta Airlines (DL) flights in December
delta_december_flights <- my_flights %>%
  filter(carrier == "DL", month == 12)

# Calculate the 95th percentile (0.95 quantile) of arr_delay
percentile_95 <- quantile(delta_december_flights$arr_delay, probs = 0.95)

cat("The 95th percentile of arrival delays for Delta Airlines (DL) flights in December is:", percentile_95, "minutes ")

``` 

6. \*Now, use `qnorm()` to find the top 5% according to the *true* normal distribution with the same mean and standard deviation as you found in Question 1. Then, explain how this compares to your answer in question 6? Why are they different or the same?
```{r}
# Probability for the top 5%
probability_top_5 <- 0.05

# Calculate the value from the true normal distribution
top_5_value <- qnorm(1 - probability_top_5, mean = arr_delay_stats$x_bar, sd = arr_delay_stats$s)

cat("The value from the true normal distribution corresponding to the top 5% is approximately:", top_5_value , "\n","The value from `qnorm` differs from the top 5% meaning deviation from a normal distribution is occuring, this could be from the high peaks found in the histogram")

```

***

## Exploratory Data Analysis

#### Research Question: Do flights that are longer in distance have more arrival delays?  
*Note: for the remaining questions, please use the full dataset `nycflights`.

7. What is the response variable and what is the predictor of interest? Specify the type of each variable: quantitative (discrete or continuous) or categorical (ordinal or nominal).

The response variable is "arrival delay"a quantitative continuous variable, which represents the amount of time a flight arrives early or late compared to its scheduled arrival time (in minutes). 
Additionally, "dep_delay" a quantitative continuous. Represents the amount of time a flight departs early or late compared to its scheduled departure time (in minutes). 

The predictors of interest could be 
Carrier: This is a categorical nominal variable representing the airline carrier responsible for the flight.
Month: This is a categorical nominal variable representing the month in which the flight occurred.
Departure Airport: This is a categorical nominal variable representing the airport from which the flight departed.
Distance: This is a quantitative continuous variable representing the distance (in miles) of the flight.
Air Time: This is a quantitative continuous variable representing the time (in minutes) the flight was in the air.


8. What are two potential confounding variables that could influence the relationship between departure delays and arrival delays. At least one should be from the dataset. Cite any sources used. Explain how the variable could be associated with the predictor of interest *and* the response.

One potential confounding variable that could influence the relationship between departure delays and arrival delays is the month of the departure. Many people might be off from work, school and might have holidays during a specif time of the year so there might be more congestion in the airports during those times. 

Another possible confounding variable that could influence the relationship between departure delays and arrival delays could be weather conditions. If the weather conditions are harsh such as storms, snow, high winds, airplanes might not be able to function as expected causing possible delays on departures and arrivals. 
https://weather.com/maps/travel/airport-delays 


9. Make a well-labeled plot (using your choice of R package), to display the relationship between the predictor of interest and the response.
```{r}

# Create the scatter plot
scatter_plot <- ggplot(nycflights, aes(x = dep_delay, y = origin)) +
  geom_point() +
  labs(
    x = "Departure Delay (minutes)",
    y = "Departure Airport"
  ) +
  ggtitle("Relationship Between Departure Delay and Departure Airport")

# Display the scatter plot
print(scatter_plot)

```

10. What do you notice about the relationship?
The relationship between the departure delays and departure airport orgin seems to be that in the airport of JFK has more departure delays and the outliers are very large compared to the other two airports. But also the majority of departure delays coming from JFK are less than the other two. 


11. Adjust the plot you made in Question 9 (use a new aesthetic) to also include one of the confounding variables you mentioned in Question 8. Comment on any patterns.
```{r}

#  line plot with color differentiation
ggplot(nycflights, aes(x = dep_delay, y = arr_delay, color = factor(month))) +
  geom_point() +
  labs(
    x = "Departure Delay (Mins)",
    y = "Arrival Delay (Mins)",
    title = "Relationship Between Departure Delay and Arrival Delay (Month)",
    color = "Month"
  ) +
  theme_dark()

```
The plot above shoes the relationship between arrival delay, departure delay and how are these values distributed by month. As previously discussed, the month has a lot to do with the delays for instance during the months of the holiday season or summer vacation. More people could be at the airports meaning more crows and more delays. As seen above the highest arrival and departure delays happen in the months of December and January which is when many people travel or come back from their vacations after the holidays. But it can also be due to the weather conditions from the area. 


