---
title: 'STAT 630: Homework 1'
author: 'Andrea Labra Orozco '
date: 'Due: September 7th, 2023 at 11:59pm'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

1. Listen to the following episode of [Stats + Stories](https://statsandstories.net/society1/getting-better-data-on-where-we-live).

    a) What is the sampling unit in the American Housing Survey?
    The sampling unit in the American Housing Survey is housing units 
    
    b) What does HUD stand for?
    HUD stands for the U.S Department of Housing and Urban Development 
    
    c) How does the Census Bureau try to reduce respondent burden?
    The Census Bureau try to reduce respondent burden by using administrative 
    data sources such as housing characteristics like year built, number of bedrooms. 
    
    d) Describe the sampling process in a few sentences _in your own words_.
    The sampling process for the American Housing Survey begins with the creation
    of a national address list compiled by the U.S. Census Bureau, which includes 
    residential addresses from various sources. 
    From this list, a representative sample of about 100,000 housing units 
    is selected every two years. 
    
    e) What would you like to know about? Write a research question that could be answered with the American Housing Survey.
    One question that could be answer with the American Housing Survey 
    How does the distribution of housing costs vary across different demographic 
    groups (e.g., income levels, age, ethnicity) in the United States? 
    
2. Install the `openintro` package, by uncommenting the following code. 

*Reminder: you only have to do this once- like installing an app on your phone. After you run this line of code, either **comment** it out using `#`, or just delete it.*

```{r}
#install.packages("openintro")
```

After installing the R package for our book, load it, i.e., open the app!

```{r}
library(openintro) # Load the package
```

Load in the `babies` dataset. Use the help file to learn more.

```{r}
data(babies) # Load the data
?babies # View the help file
```

View a summary of the dataset.
```{r}
dplyr::glimpse(babies) # Glimpse the dataset
summary(babies) # View a summary of each column (variable)
```

    a) What does each row in the table represent, i.e., what is the observational unit?
    Each row in the table represents the each observation, in this case is each 
    pregnancy that was observed in this study. 
    
    b) How many participants were in the study?
    There was 1236 participants in this study 
    
    c) All variables are coded as integers. Which variables should be recoded as *factors*? Recode these variables in the code chunk below.
    
    
```{r}
babies$parity_f <- factor(babies$parity,
                          labels = c("First Pregnancy", "More than 1 Pregnancy"))

babies$smoke_f <- factor(babies$smoke, 
                         label = c("Mother does not smoke", "Mother does smoke"))



```

    d) Create a plot to visualize one variable of your choice using `base R` functions. in the code chunk below. Make sure to add a title and relabel the x and y axes.
    
```{r}
hist(babies$age,
        main = "Age of the Mothers ", 
        xlab = "Age", 
        ylab = "frequency",
        col = "pink", 
        border = "purple"
     )

```

    e) Create a plot to visualize the same variable from part (d) using `tidyverse` functions. in the code chunk below. Make sure to add a title and relabel the x and y axes.
    
```{r}
#For next homework

```

    f) What did you learn from the plots above that you did not learn from the `summary()`? Explain.
    
    The histogram plot shows the distribution of mothers' ages, revealing that the data appears to be somewhat right-skewed, with more mothers falling in the younger age categories.
    
    g) Fill in the table below. Show the code you used in the R chunk below as well.
    

```{r}

mean_motherAge <- mean(babies$age, na.rm = TRUE)
#na.rm is to ignore N/A came from https://community.rstudio.com/t/calculate-the-mean-by-colum-when-there-is-na/152100
sd_motherAge <- sd(babies$age, na.rm = TRUE)

mean_gestation <- mean(babies$gestation, na.rm = TRUE)
sd_gestation <- sd(babies$gestation, na.rm = TRUE)

mean_bwt <- mean(babies$bwt, na.rm = TRUE)
sd_bwt <- sd(babies$bwt, na.rm = TRUE)

mean_mW <- mean(babies$weight, na.rm = TRUE)
sd_mW <- sd(babies$weight, na.rm = TRUE)

#Parity factor table 
pari <- prop.table(table(babies$parity_f))
pari

#Smoke factor table 
props <- prop.table(table(babies$smoke_f))
props

```

    
    
| Variable | mean (sd) or n(/%) |
|--------|-----------------------|
| Mother's Age | `r round(mean_motherAge,2)`  (`r round( sd_motherAge,2)`) |
| Parity |  |
|-First Pregnancy| (`r round(pari[1]*100,2)`%) |
|-More than 1 Pregnancy| (`r round(pari[2]*100,2)`%) |
| Gestation | `r round(mean_gestation,2)`  (`r round(sd_gestation)`)|
| Birth weight (oz) | `r round(mean_bwt,2)` (`r round(sd_bwt,2)`) |
| Mother's weight (lbs)| `r round(mean_mW,2)` (`r round(sd_mW,2)`) |
| Smoke status |  |
|- Mother does not Smoke| (`r round(props[1]*100,2)`%) |
|- Mother Smokes| (`r round(props[2]*100,2)`%) |


    